---
title: "Getting Started with bayescomp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bayescomp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
Standards documented in vignettes/ directory for srr compliance:
@srrstats {G5.2b} Vignette code demonstrates expected model output.
@srrstats {BS1.2b} Prior specification documented with examples below.
@srrstats {RE4.1} Vignette shows multiple output forms (summary, effects, plots).
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What you will build

By the end of this tutorial you will have:

1. Read a benchmark CSV into R
2. Fitted a Bayesian hierarchical model comparing two methods
3. Extracted a publication-ready effect estimate with credible intervals
4. Plotted the results

The whole analysis takes 15 lines of R code. You do not need to know brms,
Stan, or Bayesian statistics to follow along.

## Prerequisites

Install bayescomp and make sure cmdstanr is configured (bayescomp uses it
as the Stan backend):

```{r load}
library(bayescomp)
library(dplyr)
```

## Step 1 -- Read the data

The example data ships with the package. It compares two NEB algorithms
(CI-NEB and OCI-NEB) on 24 transition-state search problems. Each row is
one system; columns like `Calls_CINEB` and `Calls_MMF` hold the force
evaluation counts.

Read it. The columns use suffixes `_CINEB` and `_MMF` (the internal label
for OCI-NEB).

```{r read}
path <- system.file("extdata", "baker_bench.csv", package = "bayescomp")

bench_raw <- bc_read_benchmark(
  path,
  format         = "wide",
  method_suffixes = c("CINEB", "MMF"),
  system_col     = "System",
  count_col      = "Calls",
  time_col       = "Time",
  success_col    = "Term"
)

dim(bench_raw)
names(bench_raw)
```

`bc_read_benchmark()` reads the CSV, converts the system column to a factor,
and attaches metadata attributes that downstream functions use automatically.

## Step 2 -- Pivot to long format

Most statistical models expect one observation per row.  `bc_pivot_long()`
reshapes the wide data so that each system-method pair becomes its own row.

```{r pivot}
bench_long <- bc_pivot_long(
  bench_raw,
  method_pattern = "_(CINEB|MMF)$",
  method_levels  = c("CINEB", "MMF"),
  count_col      = "count",
  time_col       = "time",
  success_col    = "success",
  success_value  = "GOOD"
)

head(bench_long)
```

The `method_levels` argument controls factor ordering.  The first level
(`"CINEB"`) becomes the reference method in the model, so all treatment
effects are reported relative to CI-NEB.

You can inspect the structure to make sure it looks right:

```{r structure}
str(bench_long[, c("system_id", "method", "count", "time", "success")])
```

Expected output (abbreviated):

```
tibble [48 x ...] (S3: tbl_df/tbl/data.frame)
 $ system_id: Factor w/ 24 levels ...
 $ method   : Factor w/ 2 levels "CINEB","MMF": 1 2 1 2 ...
 $ count    : int  202 177 546 300 882 374 ...
 $ time     : num  6.06 5.66 11.13 8.23 ...
 $ success  : logi  TRUE TRUE TRUE TRUE ...
```

## Step 3 -- Fit the model

`bc_fit()` selects the response family automatically (negative binomial for
integer counts, Gamma for positive continuous times, Bernoulli for binary
success).  It sets weakly informative priors and runs four MCMC chains via
cmdstanr.

The model formula is:

    count ~ method + (1 | system_id)

This is a generalized linear mixed model with a fixed effect for method and
a random intercept per test system, accounting for the fact that some
reactions are inherently harder than others.

```{r fit, eval = FALSE}
fit_count <- bc_fit(
  bench_long,
  response  = "count",
  model_shape = TRUE,
  file      = "models/baker_count"
)
```

Setting `model_shape = TRUE` adds a shape submodel (`shape ~ method`) so
that the dispersion parameter can differ between CI-NEB and OCI-NEB.  This
is important because one method may produce more variable call counts than
the other.

The `file` argument caches the compiled and fitted model to disk, so
subsequent runs load instantly.

**Note:** Model fitting takes 2-5 minutes depending on your hardware.
The code block above uses `eval = FALSE` because MCMC sampling is too
slow for automated vignette builds.

When fitting completes, bayescomp automatically runs convergence
diagnostics and prints a summary:

```
v All convergence diagnostics passed (Rhat <= 1.01, ESS >= 400).
i NB shape (reference): 1.3 (median on response scale)
i NB shape (MMF): 2.1 [1.0, 4.5]
```

The low shape values (well below 50) confirm substantial overdispersion,
validating the choice of negative binomial over Poisson.

## Step 4 -- Summarize effects

`bc_summarize_effects()` extracts the posterior distribution of the method
effect and converts it to interpretable quantities: a multiplicative factor
and a percentage change on the response scale.

```{r effects, eval = FALSE}
effects <- bc_summarize_effects(fit_count)
print(effects)
```

Expected output:

```
# A tibble: 3 x 5
  effect_type                      median  lower  upper formatted
  <chr>                             <dbl>  <dbl>  <dbl> <chr>
1 Expected Response (Baseline)     551.   331.   917.   551.23 [330.65, 917.42]
2 Multiplicative Factor (MMF)        0.54   0.45   0.63 0.54 [0.45, 0.63]
3 Percentage Change (MMF)          -46.4  -54.6  -36.7  -46.41 [-54.62, -36.73]
```

The key result: OCI-NEB uses 46.4% fewer force/energy evaluations than
CI-NEB, with a 95% credible interval of [-54.6%, -36.7%].  Because the
entire interval is negative, there is strong evidence that OCI-NEB is
more efficient.

For a publication-ready table:

```{r table, eval = FALSE}
bc_effect_table(fit_count)
```

```
# A tibble: 3 x 3
  Effect                           Median  `95% CrI`
  <chr>                            <chr>   <chr>
1 Expected Response (Baseline)     551.23  [330.65, 917.42]
2 Multiplicative Factor (MMF)      0.54    [0.45, 0.63]
3 Percentage Change (MMF)          -46.4%  [-54.6%, -36.7%]
```

## Step 5 -- Plot conditional effects

`bc_plot_conditional()` overlays the model's predicted response on the raw
data, letting you see how well the model captures the pattern across test
systems.

```{r plot, eval = FALSE}
bc_plot_conditional(
  fit_count,
  effects      = "method",
  data         = bench_long,
  response_col = "count",
  log_y        = TRUE
)
```

The resulting plot shows the posterior mean and 95% credible band for each
method, with the raw per-system counts overlaid as jittered points.  The
log scale is appropriate here because the negative binomial model uses a
log link.

## Adding covariates

The Baker data includes `RMSD_Init_Final` (the geometric dissimilarity
between reactant and product), which is known to affect the difficulty of
a NEB calculation.  You can include this as a method-varying spline:

```{r spline, eval = FALSE}
fit_spline <- bc_fit(
  bench_long,
  response         = "count",
  spline_by_method = "RMSD_Init_Final",
  model_shape      = TRUE,
  file             = "models/baker_count_spline"
)
```

This adds `s(RMSD_Init_Final, by = method, k = 3)` to the formula, letting
each method have its own smooth relationship with the covariate.  The
conditional effects plot then shows predicted counts as a function of
RMSD, coloured by method.

```{r spline-plot, eval = FALSE}
bc_plot_conditional(
  fit_spline,
  effects      = "RMSD_Init_Final:method",
  data         = bench_long,
  response_col = "count",
  log_y        = TRUE
)
```

## Using a targets pipeline

For reproducible analyses, bayescomp integrates with the targets package.
Place the following in your `_targets.R` file:

```{r targets, eval = FALSE}
library(targets)
library(bayescomp)

list(
  bc_tar_pipeline(
    data_path       = "inst/extdata/baker_bench.csv",
    method_suffixes = c("CINEB", "MMF"),
    method_levels   = c("CINEB", "MMF"),
    response        = "count",
    system_col      = "System",
    count_col       = "Calls",
    time_col        = "Time",
    success_col     = "Term",
    model_shape     = FALSE,
    model_file      = "data/models/baker_count"
  )
)
```

Run `targets::tar_make()` to execute the pipeline.  Subsequent runs skip
steps whose inputs have not changed.

## Summary

The typical bayescomp workflow is:

1. `bc_read_benchmark()` -- load your CSV data
2. `bc_pivot_long()` -- reshape from wide to long format
3. `bc_fit()` -- fit a Bayesian hierarchical model
4. `bc_summarize_effects()` or `bc_effect_table()` -- extract treatment effects
5. `bc_plot_conditional()` -- visualize the model predictions

For diagnostics, model selection, and multi-method comparisons, see the
companion vignettes:

- `vignette("neb-case-study")` -- full NEB analysis with diagnostics
- `vignette("dimer-case-study")` -- multi-method comparison (>2 methods)
- `vignette("model-selection")` -- family selection and prior sensitivity
