---
title: "Multi-Method Comparison with Dimer Rotation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Method Comparison with Dimer Rotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette analyses the dimer rotation benchmark, which compares four
optimization methods for the initial dimer rotation step in transition-state
searches.  The data set has three response variables (force/energy call
counts, wall-clock time, and binary success), making it a good test case
for `bc_fit_suite()`.

It also demonstrates two features that go beyond the basic `bc_fit()`
workflow:

1. `bc_fit_design()` for factorial designs with interaction terms
2. `bc_pairwise_contrasts()` for all-pairs comparisons when there are more
   than two methods

## Data

The dimer rotation benchmark ships with bayescomp.  Unlike the Baker data,
it is already in long format: one row per system-method observation.

```{r load}
library(bayescomp)
library(dplyr)
```

```{r read}
path <- system.file("extdata", "idrot_bench.csv", package = "bayescomp")

bench <- bc_read_benchmark(
  path,
  format     = "long",
  system_col = "system_id"
)

head(bench)
```

The columns are:

- `system_id` -- identifier combining molecule index and spin state
- `method` -- one of `cg_no`, `cg_yes`, `lbfgs_no`, `lbfgs_yes`
  (conjugate-gradient or L-BFGS optimizer, with or without rotation removal)
- `pes_calls` -- number of potential energy surface evaluations (count)
- `tot_time` -- wall-clock time in seconds (positive continuous)
- `success` -- logical success flag derived from `termination_status`
- `mol_id` -- molecule index (grouping variable)
- `spin` -- spin state (doublets or singlets)

```{r levels}
bench$method <- factor(
  bench$method,
  levels = c("cg_no", "cg_yes", "lbfgs_no", "lbfgs_yes")
)

table(bench$method)
```

The reference level is `cg_no` (conjugate gradient without rotation
removal), against which all other methods are compared.

## Validate the data

```{r validate}
bc_validate(
  bench,
  count_col  = "pes_calls",
  method_col = "method",
  system_col = "system_id"
)
```

## Simple model with bc_fit()

For a first pass, you can use `bc_fit()` with the count response:

```{r simple-fit, eval = FALSE}
fit_simple <- bc_fit(
  bench,
  response    = "pes_calls",
  method_col  = "method",
  system_col  = "system_id",
  model_shape = TRUE,
  file        = "models/idrot_simple"
)

bc_summarize_effects(fit_simple)
```

This gives you percentage change estimates for each method relative to
`cg_no`.  However, the four methods form a 2x2 factorial design
(optimizer x rotation removal), and `bc_fit()` treats them as four
independent levels.  To test the interaction between optimizer choice and
rotation removal, you need `bc_fit_design()`.

## Factorial design with bc_fit_design()

First, create the factorial columns from the method labels:

```{r factors}
bench <- bench %>%
  mutate(
    optimizer = factor(
      ifelse(grepl("^cg", method), "CG", "LBFGS"),
      levels = c("CG", "LBFGS")
    ),
    rotation_removal = factor(
      ifelse(grepl("_yes$", method), "yes", "no"),
      levels = c("no", "yes")
    )
  )
```

Now specify the interaction formula using `brms::bf()`:

```{r design-fit, eval = FALSE}
library(brms)

design_formula <- bf(
  pes_calls ~ optimizer * rotation_removal + (1 | system_id)
)

fit_design <- bc_fit_design(
  bench,
  formula = design_formula,
  family  = negbinomial(link = "log"),
  file    = "models/idrot_design"
)
```

The interaction term `optimizer:rotation_removal` tells us whether the
benefit of rotation removal depends on which optimizer you use (or vice
versa).

```{r design-effects, eval = FALSE}
bc_summarize_effects(fit_design)
```

## Nested random effects

The dimer data has a hierarchical grouping structure: each molecule
(`mol_id`) is tested in two spin states (`spin`), creating the
`system_id = mol_id:spin` compound key.  You can model this explicitly:

```{r nested, eval = FALSE}
nested_formula <- bf(
  pes_calls ~ optimizer * rotation_removal + (1 | mol_id) + (1 | mol_id:spin)
)

fit_nested <- bc_fit_design(
  bench,
  formula = nested_formula,
  family  = negbinomial(link = "log"),
  file    = "models/idrot_nested"
)
```

This partitions variance into molecule-level and spin-state-level
components.  Use `bc_loo_compare()` to check whether the nested structure
improves prediction (see the model-selection vignette).

## Fitting a suite of models

When your data includes count, time, and success columns, `bc_fit_suite()`
fits a separate model for each, using the appropriate family automatically:

- Negative binomial for `pes_calls` (counts)
- Gamma for `tot_time` (positive continuous)
- Bernoulli for `success` (binary)

```{r suite, eval = FALSE}
# Ensure the success column is properly coded
bench$success_bin <- as.integer(bench$success == "True" | bench$success == TRUE)

suite <- bc_fit_suite(
  bench,
  count_col   = "pes_calls",
  time_col    = "tot_time",
  success_col = "success_bin",
  method_col  = "method",
  system_col  = "system_id",
  file_prefix = "models/idrot_suite"
)

names(suite)
```

Expected output:

```
[1] "count" "time" "success"
```

Each element is a `brmsfit` object.  You can extract effects from each:

```{r suite-effects, eval = FALSE}
lapply(suite, bc_summarize_effects)
```

## Pairwise contrasts

With four methods, there are six pairwise comparisons.
`bc_pairwise_contrasts()` uses the marginaleffects package to compute all
of them from a single fitted model.

```{r contrasts, eval = FALSE}
contrasts <- bc_pairwise_contrasts(suite$count, method_col = "method")
print(contrasts)
```

Expected output (values are illustrative):

```
# A tibble: 6 x 5
  contrast             estimate  lower  upper formatted
  <chr>                   <dbl>  <dbl>  <dbl> <chr>
1 cg_yes - cg_no         -85.2 -142.3  -28.1 -85.20 [-142.30, -28.10]
2 lbfgs_no - cg_no       -62.4 -118.7   -6.1 -62.40 [-118.70, -6.10]
3 lbfgs_yes - cg_no     -120.6 -176.9  -64.3 -120.60 [-176.90, -64.30]
4 lbfgs_no - cg_yes       22.8  -33.5   79.1 22.80 [-33.50, 79.10]
5 lbfgs_yes - cg_yes     -35.4  -91.7   20.9 -35.40 [-91.70, 20.90]
6 lbfgs_yes - lbfgs_no   -58.2 -114.5   -1.9 -58.20 [-114.50, -1.90]
```

Contrasts whose credible intervals exclude zero provide strong evidence of
a difference.  Contrasts that span zero are inconclusive at the chosen
credibility level.

## Plotting contrasts

The contrast plot is a horizontal pointrange chart with a vertical dashed
line at zero:

```{r plot-contrasts, eval = FALSE}
bc_plot_contrasts(contrasts)
```

## When to use bc_fit() vs bc_fit_design()

| Situation                                 | Function          |
|-------------------------------------------|-------------------|
| Two methods, simple comparison            | `bc_fit()`        |
| Two methods with spline covariate         | `bc_fit()`        |
| Three or more methods, no factorial       | `bc_fit()`        |
| Factorial design (optimizer x setting)    | `bc_fit_design()` |
| Custom random effects structure           | `bc_fit_design()` |
| Interaction terms                         | `bc_fit_design()` |

`bc_fit()` constructs the formula for you from column names.
`bc_fit_design()` gives you full control over the brms formula while still
providing bayescomp defaults for priors, diagnostics, and backends.

## Summary

Key bayescomp functions demonstrated in this vignette:

- `bc_read_benchmark()` with `format = "long"` for pre-pivoted data
- `bc_validate()` to check data structure before fitting
- `bc_fit_design()` with `brms::bf()` for factorial interaction models
- `bc_fit_suite()` to fit count, time, and success models in one call
- `bc_pairwise_contrasts()` for all-pairs method comparisons
- `bc_plot_contrasts()` for visualizing pairwise differences
