---
title: "NEB Method Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NEB Method Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette reproduces the full Bayesian analysis from the OCI-NEB paper,
comparing the OCI-NEB method (labelled `MMF` in the data) against the
standard CI-NEB method on the Baker benchmark set of 24 transition-state
search problems.

The analysis covers:

- Data preparation with RMSD spline covariate
- Negative binomial model with shape submodel and censoring
- All six diagnostic plots
- Effect extraction and publication-ready tables

## Data preparation

```{r load}
library(bayescomp)
library(dplyr)
```

```{r read}
path <- system.file("extdata", "baker_bench.csv", package = "bayescomp")

bench_raw <- bc_read_benchmark(
  path,
  format          = "wide",
  method_suffixes = c("CINEB", "MMF"),
  system_col      = "System",
  count_col       = "Calls",
  time_col        = "Time",
  success_col     = "Term"
)
```

Pivot to long format, adding a binary censoring indicator for systems that
hit the maximum iteration count.  The `cens_value` argument flags any
observation whose termination status matches the given string.

```{r pivot}
bench_long <- bc_pivot_long(
  bench_raw,
  method_pattern = "_(CINEB|MMF)$",
  method_levels  = c("CINEB", "MMF"),
  count_col      = "count",
  time_col       = "time",
  success_col    = "success",
  success_value  = "GOOD",
  cens_value     = "BAD_MAX_ITERATIONS"
)
```

The RMSD between the initial and final NEB images measures the geometric
difficulty of the reaction path.  This column was already pivoted into the
long data as `RMSD_Init_Final`.  Verify it is present:

```{r rmsd-check}
"RMSD_Init_Final" %in% names(bench_long)
```

## Model fitting

The full model specification includes:

- A fixed effect for `method` (CI-NEB vs OCI-NEB)
- A method-varying thin plate spline on `RMSD_Init_Final`
- A random intercept per test system
- A shape submodel (`shape ~ method`) so overdispersion can differ
- Right-censoring via the `censored` column

```{r fit, eval = FALSE}
fit <- bc_fit(
  bench_long,
  response         = "count",
  spline_by_method = "RMSD_Init_Final",
  model_shape      = TRUE,
  cens_col         = "censored",
  chains           = 4,
  iter             = 4000,
  warmup           = 1000,
  seed             = 1995,
  file             = "models/baker_full"
)
```

## Diagnostic 1 -- Convergence check

`bc_check_convergence()` examines Rhat and effective sample size (ESS) for
every parameter.  The thresholds follow current best practice: Rhat below
1.01 and bulk ESS above 400.

```{r convergence, eval = FALSE}
conv <- bc_check_convergence(fit)
conv$passed
```

Expected output:

```
v All convergence diagnostics passed (Rhat <= 1.01, ESS >= 400).
[1] TRUE
```

If convergence fails, increase `iter` or `adapt_delta` (already set to 0.99
by default).  Inspect the problematic parameters:

```{r conv-details, eval = FALSE}
conv$problems
conv$max_rhat
conv$min_ess
```

## Diagnostic 2 -- LOO-CV

Leave-one-out cross-validation via Pareto-smoothed importance sampling
detects observations that the model struggles to predict.

```{r loo, eval = FALSE}
loo_result <- bc_loo(fit)
```

Expected output:

```
i LOO-CV: elpd = -283.2
i Pareto k: 48 good (k <= 0.5), 2 marginal, 0 bad (k > 0.7)
```

No observations with Pareto k above 0.7 means the LOO approximation is
reliable and no refitting is needed.

## Diagnostic 3 -- Posterior predictive checks

Posterior predictive checks compare the observed data distribution against
data simulated from the fitted model.  If the model is well-specified,
simulated data should look similar to the observed data.

```{r pp-check, eval = FALSE}
pp <- bc_pp_check(fit)
pp$density
pp$grouped
```

The density overlay (`pp$density`) shows the observed density (dark line)
against 50 simulated densities (light lines).  Good overlap indicates
adequate fit.

The grouped check (`pp$grouped`) stratifies by method, letting you verify
that the model captures both CI-NEB and OCI-NEB distributions separately.

## Diagnostic 4 -- LOO-PIT QQ plot

The LOO-PIT (Probability Integral Transform) plot assesses calibration.
If the model is well-calibrated, the LOO-PIT values follow a uniform
distribution, and the QQ plot falls along the diagonal.

```{r loo-pit, eval = FALSE}
bc_plot_loo_pit(fit, loo_result, y = bench_long$count)
```

Deviations from the diagonal indicate systematic miscalibration.  A
concave pattern (points above the line on the left, below on the right)
suggests the model is overdispersed relative to the data.

## Diagnostic 5 -- Posterior predictive density

The standalone PP density plot uses `theme_bayescomp()` styling:

```{r pp-density, eval = FALSE}
bc_plot_pp(fit, ndraws = 50, type = "dens_overlay")
```

## Diagnostic 6 -- Shape parameter comparison

For negative binomial models with `model_shape = TRUE`, the shape
parameter captures overdispersion.  Lower shape means more overdispersion
(more variance relative to the mean).

```{r shape, eval = FALSE}
bc_plot_shape(fit)
```

This density plot shows the posterior distribution of the log-scale shape
parameter for each method.  If the two densities are well-separated, the
methods have genuinely different variance structures.

## Effect extraction

```{r effects, eval = FALSE}
effects <- bc_summarize_effects(fit)
print(effects)
```

Expected output:

```
# A tibble: 3 x 5
  effect_type                      median  lower  upper formatted
  <chr>                             <dbl>  <dbl>  <dbl> <chr>
1 Expected Response (Baseline)     551.   331.   917.   551.23 [330.65, 917.42]
2 Multiplicative Factor (MMF)        0.54   0.45   0.63 0.54 [0.45, 0.63]
3 Percentage Change (MMF)          -46.4  -54.6  -36.7  -46.41 [-54.62, -36.73]
```

The headline result: OCI-NEB requires 46.4% fewer force/energy evaluations
than CI-NEB (95% CrI: [-54.6%, -36.7%]).

## Publication table

```{r table, eval = FALSE}
bc_effect_table(fit)
```

```
# A tibble: 3 x 3
  Effect                           Median  `95% CrI`
  <chr>                            <chr>   <chr>
1 Expected Response (Baseline)     551.23  [330.65, 917.42]
2 Multiplicative Factor (MMF)      0.54    [0.45, 0.63]
3 Percentage Change (MMF)          -46.4%  [-54.6%, -36.7%]
```

## Forest plot

The forest plot provides a visual summary of the percentage change with
credible intervals.  A vertical dashed line at zero marks no effect.

```{r forest, eval = FALSE}
bc_plot_forest(fit)
```

## Conditional effects plot

The conditional effects plot shows predicted counts as a function of the
RMSD covariate, with separate curves for each method and raw data points
overlaid.

```{r conditional, eval = FALSE}
bc_plot_conditional(
  fit,
  effects      = "RMSD_Init_Final:method",
  data         = bench_long,
  response_col = "count",
  log_y        = TRUE
)
```

Both methods show increasing call counts with RMSD (harder reactions need
more iterations), but OCI-NEB's curve is consistently lower.  The gap
widens for high-RMSD systems, where OCI-NEB's advantage is most pronounced.

## Time analysis

You can also model wall-clock time using a Gamma family:

```{r time, eval = FALSE}
fit_time <- bc_fit(
  bench_long,
  response  = "time",
  family    = brms::brmsfamily("Gamma", link = "log"),
  file      = "models/baker_time"
)

bc_summarize_effects(fit_time)
```

The Gamma family is appropriate for positive continuous data with right
skew, which is typical of computation times.

## Summary of diagnostics

| Diagnostic                  | What it checks                | Pass criterion               |
|-----------------------------|-------------------------------|------------------------------|
| `bc_check_convergence()`    | MCMC mixing                   | Rhat <= 1.01, ESS >= 400    |
| `bc_loo()`                  | Predictive accuracy           | No Pareto k > 0.7           |
| `bc_pp_check()`             | Distributional fit            | Visual overlap               |
| `bc_plot_loo_pit()`         | Calibration                   | Points near diagonal         |
| `bc_plot_pp()`              | Density overlap               | Simulated matches observed   |
| `bc_plot_shape()`           | Dispersion structure          | Shape separation if expected |
